Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-17 doc/ApTranscriptome_manuscript.Rmd
completed Discussion draft

For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that species (*NotResp*). For the *Bimodal* group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

@@

To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

@@

As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A. picea* species would have a broader thermal range of stable gene expression than *A. carolinensis*. To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A. carolinensis* than *A. picea*. We performed the same analysis as with the *Intermediate* transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one species had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.

Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression. We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes. We then tested if the temperature at which the *High* or *Low* group differed among the two species.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("../ms.RData")
library(data.table)
library(pander)
library(ggplot2)
library(topGO)
library(RColorBrewer)
```

@@

We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR. Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. Thus, at a minimum ... of genes are thermally-responsive. However, the closest related ant species with a sequenced genome has only ... genes. If we use this as

The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the *Low* category (Fig. 1). For *A. carolinensis*, this was followed by *Intermediate*, *High* and then *Bimodal*, whereas for *A. picea* the second largest category was *Bimodal* followed by *High* and finally *Intermediate* (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for *High* and *Low* genes, and the variance in expression for the *Intermediate* and *Bimodal* genes. The *High* genes in *A. carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at *Low* temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2). For *Bimodal* expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`). In contrast, *Intermediate* genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`).

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}

@@

We looked for evidence of genetic assimilation by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, for *Low* genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for *High* genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and *Intermediate* genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For *Intermediate* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, *Intermediate*, in *A. carolinensis* to having increased expression in *Low* temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For *Intermediate* genes in *A. carolinensis* that shifted *Bimodal* in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

@@

As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

@@

- many genes thermally responsive
- thermal reaction norms differ between species
  - *A. carolinensis* limited future potential to adapt to heat stress
- more genes activated in response to cold shock --> greater adaptability to cold
- differences among species 
  - *A. carolinensis* tolerates thermal stress by shutting down biological processes
  - in *A. picea* the *Intermediate* genes from *A. carolinensis* are over-represented in *Low* and *Intermediate* categories
  - potential explanation is that *A. carolinensis* tolerates thermal stress by shutting-down metabolism and relies on other mechanisms (behavior, protein stability) whereas *A. picea* activates protective genes which requires metabolism...
- comparison of gene terms with other species (Sara?)

- regression approach allows us to detect patterns not evident from ANOVA design

-------------------------------------

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [@krebs1997; @gaitainespitia2013], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively. We found that at least 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress. The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A. carolinensis* and more up-regulation of gene expression in *A. picea* in response to thermal stress (Fig. 2).

Given the differences in the number of thermally-responsive genes between the two species (Table 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species, with greater variation between species at low temperatures (Fig. 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures. This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. Moreover, in an experiment with 94 species of *Drosophila*, Kellerman et al [-@kellerman2012] showed that limited genetic variation among species (e.g. phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits. Thus, our results indicate that, when coupled with phenotypic data, transcriptomic studies can provide insight on adapatability. For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal].

Depending on the frequency with which organisms experience stressful conditions, the reaction norms of protective genes may differ considerably. Organisms that frequently experience stress may have high basal levels of expression with little change, while organisms that less frequently experiences stress would have lower basal levels with large changes in response to environmental stress [@cite...]. For *Aphaenogaster*, we predicted that genes up-regulated to cold or hot temperatures in one species might be expressed at higher levels near optimal growth temperatures in the other species to confer greater basal stress tolerance. This prediction was not supported, as we found greater expression near optimal temperatures in *A. carolinensis* for *High* genes in *A. picea*, as predicted, but also for *Low* genes in *A. carolinensis* and *Intermediate* genes in both species. As *A. carolinensis* is from a warmer environment, a likely explanation for this is that this species simply has a higher metabolic rate and thus greater gene expression across it's thermal range. The observation that there are more *Intermediate* genes, which are shut-down under thermal stress, in *A. carolinensis* supports this explanation. ...cite some coral literature....

@@

### Molecular toolkit of thermal tolerance

REANALYZING...

@@

As with all *de novo* transcriptomes, our work has some important caveats. First, given that all sequenced ant genomes to date have about 18,000 genes [@smith...], our transcriptome is highly fragmented. Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, would considerably resolve this problem. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may only have small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change. Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@cite...]. In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@cite...].

@@

In this work, we have completed the first transcriptome for a temperate ant species, *Aphaenogaster picea* and *A. carolinensis*, to date. We found that a large proportion of the transcriptome is thermally-responsive. Transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures. ...differences among species...*A. carolinensis* greater organismal tolerance and shuts down metabolism under thermal stress...*A. picea* less tolerant and activates protective proteins requiring metabolism... These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

>>>>

For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

@@

To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category *Not Responsive*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

@@

As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A. picea* species would have a broader thermal range of stable gene expression than *A. carolinensis*. To test this with our data, for each **Intermediate** expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Conversely, we predicted that the thermal sensitivity of **Bimodal** expressed genes would be greater in *A. carolinensis* than *A. picea*. We performed the same analysis as with the **Intermediate** transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity.

Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression. We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes. We then tested if the temperature at which the **High** or **Low** group differed among the two species.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("../ms.RData")
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(topGO)
```

@@

We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR. Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive. However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012], and assuming the number is similar for *Aphaenogaster*, this suggests that up to 50% of the genome could be thermally-responsive. The reaction norms for each responsive gene can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. These differences were not simply of magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes, and the variance in expression for the **Intermediate** and **Bimodal** genes. The **High** genes in *A. carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at **Low** temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2). For **Bimodal** expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`). In contrast, **Intermediate** genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`).

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}

@@

We looked for evidence of genetic assimilation by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, for **Low** genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for **High** genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A. carolinensis* to having increased expression in **Low** temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For **Intermediate** genes in *A. carolinensis* that shifted **Bimodal** in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

@@

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013]. We found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.

The patterns of gene expression between species differed substantially, with more down-regulation in *A. carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A. picea* in response to thermal extremes (**High** and **Low** genes; Fig. 2). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are up-regulated at cold temperatures (**Low**) in *A. picea* but shut-down **Intermediate** by *A. carolinensis* include metabolism, transport and regulation GO clusters. Thus, *A. carolinensis* shuts-down metabolism to a lesser extent than *A. picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis*, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically-enduced resistance.

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig. 1, Fig. 2). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [cite...kingsolver, chown]. This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. Moreover, in an experiment with 94 species of *Drosophila*, Kellerman et al [-@kellerman2012] showed that limited genetic variation among species (e.g. phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits. Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential. For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal2012].

For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed  [@barshis2013]. However, in contrast to these expectations, we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both spcies. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.

@@

@@

As with all *de novo* transcriptomes, our work has some important caveats. First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, would considerably resolve this problem. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may only have small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change. Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@cite...]. In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@cite...].

@@

In this work, we have completed the first transcriptome for a temperate ant species, *Aphaenogaster picea* and *A. carolinensis*, to date. Using a novel regression approach, we found that a large proportion of the transcriptome is thermally-responsive. These data indicate that the two species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock and other protective molecules, while *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolernace. For both species, transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures. These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.
